Studying the Performance of the Jellyfish Optimiser for the Application of Projection Pursuit

### 2024-04-12 optim.tex
smoothness measure
Update of the smoothness measure part with equations

\begin{keyword}
    projection pursuit \sep optimization \sep jellyfish
optimiser \sep data visualisation \sep 
    high-dimensional data
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, interior hidden, boxrule=0pt, frame hidden, enhanced, breakable, borderline west={3pt}{0pt}{shadecolor}]}{\end{tcolorbox}}\fi

@@

@@

The jellyfish optimiser (JSO) mimics the natural movements of jellyfish,
which include passive and active motions driven by ocean currents and
their swimming patterns, respectively. In the context of optimization,
these movements are abstracted to explore the search space in a way that
balances exploration (searching new areas) and exploitation (focusing on
promising areas). The algorithm aims to find the optimal solution by
adapting the jellyfish's behavior to navigate towards the best solution
over iterations \citep{chou_novel_2021}.

To understand what the jellyfish optimizer is doing in the context of
Projection Pursuit, we first start with a current projection (the
starting point). Then we evaluate this projection using an index
function, which tells us how good the current projection is. We then
move the projection in a direction determined by the `best jelly' and
random factors, influenced by how far along we are in the optimization
process (the trial \(i\) and \texttt{max.tries}). Occasionally, we might
explore completely new directions like a jellyfish might with ocean
currents. Then, we compare new potential projections to our current one.
If they're better, we adopt them; if not, we stick with our current
projection. This process continues and iteratively improves the
projection, until we reach the maximum number of trials.

\begin{tcolorbox}[enhanced jigsaw, coltitle=black, titlerule=0mm, opacitybacktitle=0.6, leftrule=.75mm, left=2mm, bottomtitle=1mm, bottomrule=.15mm, toprule=.15mm, colframe=quarto-callout-note-color-frame, breakable, colbacktitle=quarto-callout-note-color!10!white, colback=white, rightrule=.15mm, opacityback=0, toptitle=1mm, title={Algorithm: Jellyfish Optimizer Pseudo Code}, arc=.35mm]

@@

\begin{quote}
\end{quote}

@@

If we evaluate the index function at some random points (like the random
initialization of the jellyfish optimizer), then we can interpret these
random index values as a random field, indexed by a space parameter: the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process. There exist R packages to fit the hyperparameters of a
Gaussian process, for example ExaGeoStatR
(https://github.com/ecrc/exageostatR), which is an R wrapper for the
ExaGeoStat library (https://arxiv.org/pdf/1708.02835.pdf).

How can we define a measure of smoothness from this? A Gaussian process
is fully defined by its mean function and covariance function. The way
the covariance function is defined is where smoothness comes into play:
if an index is very smooth, then two close projection angles should
produce close index values (strong correlation); by contrast, if an
index is not smooth, then two close projection angles might give very
different index values (fast decay of correlations with respect to
distance between angles).

Luckily, the most popular class of covariance functions, known as Matérn
covariance functions (see https://arxiv.org/pdf/1708.02835.pdf on page
3) has one parameter, Theta3, which controls the decay of the covariance
function, which in turns controls the smoothness of the random field.
The ExaGeoStat paper calls it smoothness parameter: the higher Theta3,
the smoother the random field. In our context, the higher Theta3, the
smoother the index function.

This suggests the following approach:

\begin{itemize}
\tightlist
\item
  Evaluate the index function at some random projection angle values
  (our initial jellies),
\item
  Use this sample to fit a Gaussian process with Matérn covariance using
  ExaGeoStatR,
\item
  Use the fitted Theta3 parameter value as our definition of smoothness
  of the index function: the higher Theta3, the smoother the index.
\end{itemize}

@@

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-4-1.pdf}

@@

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-6-1.pdf}

>>>>

\begin{keyword}
    projection pursuit \sep optimization \sep jellyfish
optimiser \sep data visualisation \sep 
    high-dimensional data
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, boxrule=0pt, interior hidden, enhanced, breakable]}{\end{tcolorbox}}\fi

@@

\begin{verbatim}
Warning: package 'ggplot2' was built under R version 4.3.2
\end{verbatim}

\begin{verbatim}
Warning: package 'tidyr' was built under R version 4.3.2
\end{verbatim}

@@

The jellyfish optimiser (JSO) mimics the natural movements of jellyfish,
which include passive and active motions driven by ocean currents and
their swimming patterns, respectively. In the context of optimization,
these movements are abstracted to explore the search space in a way that
balances exploration (searching new areas) and exploitation (focusing on
promising areas). The algorithm aims to find the optimal solution by
adapting the jellyfish's behaviour to navigate towards the best solution
over iterations \citep{chou_novel_2021}.

To understand what the jellyfish optimizer is doing in the context of
Projection Pursuit, we first start with a current projection (the
starting point). Then, we evaluate this projection using an index
function, which tells us how good the current projection is. We then
move the projection in a direction determined by the `best jelly' and
random factors, influenced by how far along we are in the optimization
process (the trial \(i\) and \texttt{max.tries}). Occasionally, we might
explore completely new directions like a jellyfish might with ocean
currents. Then, we compare new potential projections to our current one.
If they're better, we adopt them; if not, we stick with our current
projection. This process continues and iteratively improves the
projection, until we reach the maximum number of trials.

\begin{tcolorbox}[enhanced jigsaw, title={Algorithm: Jellyfish Optimizer Pseudo Code}, bottomrule=.15mm, colframe=quarto-callout-note-color-frame, bottomtitle=1mm, titlerule=0mm, toprule=.15mm, colback=white, colbacktitle=quarto-callout-note-color!10!white, opacityback=0, opacitybacktitle=0.6, leftrule=.75mm, toptitle=1mm, coltitle=black, breakable, arc=.35mm, left=2mm, rightrule=.15mm]

@@

@@

If we evaluate the index function at some random points (like the random
initialization of the jellyfish optimizer), then we can interpret these
random index values as a random field, indexed by a space parameter: the
random projection angle. This analogy suggests to use this random
training sample to fit a spatial model, a simple one being a (spatial)
Gaussian process.

How can we define a measure of smoothness from this? The distribution of
a Gaussian process is fully determined by its mean function and
covariance function. The way the covariance function is defined is where
smoothness comes into play: if an index is very smooth, then two close
projection angles should produce close index values (strong
correlation); by contrast, if an index is not smooth, then two close
projection angles might give very different index values (fast decay of
correlations with respect to distance between angles).

Popular covariance functions are parametric positive semi-definite
functions, some of which have a parameter to capture the smoothness of
the Gaussian field. In particular, consider the Matérn class of
covariance functions, defined by

\[
K(u):=\frac{(\sqrt{2\nu}u)^{\nu}}{\Gamma(\nu)2^{\nu-1}}\mathcal{K}_{\nu}(\sqrt{2\nu}u)
\]

where \(\nu>0\) is the smoothness parameter and where
\(\mathcal{K}_\nu\) is the modified Bessel function. The Matérn
covariance function can be expressed analytically when \(\nu\) is a
half-integer, the most popular values in the literature being \(1/2\),
\(3/2\) and \(5/2\) . The parameter \(\nu\), called smoothness
parameter, controls the decay of the covariance function. As such, it is
an appropriate measure of smoothness of a random field.

In our context, we suggest to use this parameter as a measure of the
smoothness of the index function by fitting a Gaussian process prior
with Matérn covariance on a dataset generated by random evaluations of
the index function, as in the initial stage of the jellyfish random
search. There exist several R packages, such as GpGp or ExaGeoStatR, to
fit the hyperparameters of a GP covariance function on data. In this
project, we make use of the GpGp package.

The fitted value \(\nu>0\) can be interpreted as follows: the higher
\(\nu\), the smoother the index function.

@@

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

{\centering \includegraphics{optim_files/figure-pdf/unnamed-chunk-3-1.pdf}
